---
title: "Exploring the Iris Dataset: A Visual Analysis of Sepal and Petal Characteristics"
subtitle: ""
excerpt: "In this analysis, we use ggplot2 in R to visually explore how the sepal and petal dimensions vary across species. Through various plots, including scatter plots, box plots, and histograms, we aim to identify trends, correlations, and the distribution of these measurements, providing a deeper understanding of the iris flowers' physical characteristics and how they differ between species."
date: 2024-11-12
author: "Kelvin Kiprono"
draft: false
tags:
- Statistical Data Analysis
categories:
- ggplot2

---
To explore the iris dataset using ggplot2, you can create a variety of visualizations that help in understanding the relationships between different variables, such as scatter plots, box plots, and pair plots. 

Here’s how you can get started:

## Loading required package and data 
```{r}
library(ggplot2)
data("iris")
head(iris)
str(iris)
```

```{r}
summary(iris)

```
## Scatter Plot: Sepal Length vs. Sepal Width
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  labs(title = "Scatter Plot: Sepal Length vs Sepal Width",
       x = "Sepal Length",
       y = "Sepal Width")
```
## Box Plot: Sepal Length by Species 

A box plot can be useful to examine the distribution of Sepal.Length across the different species of iris.

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot() +
  geom_point()+
  geom_jitter()+
  labs(title = "Box Plot: Sepal Length by Species",
       x = "Species",
       y = "Sepal Length")
```
## Histogram: Petal Length Distribution

A histogram can help visualize the distribution of Petal.Length across the dataset. You can also use the fill argument to color the bars by species.
```{r}
ggplot(iris, aes(x = Petal.Length, fill = Species)) +
  geom_histogram(binwidth = 0.2, alpha = 0.7, position = "dodge") +
  labs(title = "Histogram: Petal Length Distribution",
       x = "Petal Length",
       y = "Frequency")
```
## Pair Plot (Pairwise Relationships Between Variables)

A pair plot (also known as a pairs plot) visualizes pairwise relationships between multiple variables in the dataset. You can use GGally::ggpairs() for this, but here's a basic ggplot solution for just two variables.

```{r}
library(GGally)
ggpairs(iris, aes(color = Species))
```
## Density Plot: Sepal Width by Species

A density plot can show the distribution of Sepal.Width across different species. This is an alternative to the histogram.

```{r}
ggplot(iris, aes(x = Sepal.Width, fill = Species)) +
  geom_density(alpha = 0.7) +
  labs(title = "Density Plot: Sepal Width by Species",
       x = "Sepal Width",
       y = "Density")
```
## Faceted Plot  

Faceting allows you to create multiple subplots based on a factor (e.g., Species). Here's how you can create separate scatter plots for each species:
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(aes(color = Species)) +
  facet_wrap(~ Species) +
  labs(title = "Faceted Scatter Plot: Sepal Length vs Sepal Width")
```
## Correlation Heatmap

To visualize the correlation between numerical variables in the iris dataset, you can use a correlation heatmap.
```{r}
library(corrplot)
# Correlation matrix
cor_matrix <- cor(iris[, 1:4])
corrplot(cor_matrix, method = "circle", type = "upper", tl.col = "black", tl.srt = 45)
```
With ggplot2, you can easily generate a variety of insightful plots to explore the iris dataset and understand the relationships between different variables.
